---
title: "Workout Activity Prediction with Weight Lifting Exercises Dataset"

output: 
  html_document
    
---

activityPrediction.Rmd:

###     <br><br>

###     A Project for the Coursera John Hopkins Machine Learning Class

###     <br><br>

###     Prepared by Marcel Merchat
###     February 1, 2016

###     <br><br>

##      Unilateral Dumbell Biceps Curl Exercize
###     An Analysis of a Human Activity Recognition Data Set (HAR)

###     <br><br>

##      Overview: 

###     We will build a prediction model for recognizing correct weight lifting
###     techniques using accelerometer data of the Human Activity
###     Recognition Data Set (HAR) documented by Velloso et al. in a 2013 study.
###     The performance of unilateral biceps curl exercise with dumbbells of
###     six male subjects with little weight lifting experience was recorded
###     using four accelerometers including three worn on the arm, glove, and
###     belt and another on the dumbbell.

###     <br><br>

###     The subjects performed one set of 10 repetitions of the biceps burl in
###     five different ways using a relatively light dumbbell (1.25kg). They 
###     were graded as they were perfoming the curls as follows:

###     <br><br>

###     Class A -- Correctly according to the specification 
###     Class B -- Throwing the elbows to the front
###     Class C -- Lifting the dumbbell halfway
###     Class D -- Lowering the dumbbell halfway
###     Class E -- Throwing the hips to the front

###     <br><br>

###     While Class A corresponds to the correct way of doing exercise, all of
###     the other classes correspond to common mistakes. 

###     <br><br>

#       Model Building 

###     <br><br>

###     We tried the random forest rf method and the regression method rpart
###     within the caret train function. The rf method was chosen because it
###     had the best accuracy of about 97% as shown in Figure-1. The accuracy of
###     the rpart method was only about 48.3%.

###     <br><br>

###     The paper on the HAR study by Velloso et al. indicates that
###     they used seventeen variables and calculated statistics to predict the
###     same exercise classifications that we are concerned with. In contrast,
###     we only use the stated data in the file for our predictions without  
###     statistics, but we are influenced by their list of seventeen 
###     parameters. We reduced the number of clean data columns from 60 to 38
###     keeping many columns that pertained to list of seventeen statistics.

###     <br><br>

#       Cross Validation 

###     The models were built using a Build data frame that was divided into
###     training and testing parts. Finally for the project Quiz, the test data
###     frame was set equal to the Quiz data which was used as Validation data.

###     <br><br>

```{r processing, results='hide', echo = FALSE, message=F, warning=F}

##library(dplyr)
library(lattice)
library(psych)
library(ggplot2)
library(caret)
library(randomForest)
library(AppliedPredictiveModeling)
library(xtable)
library(gridExtra)

#overwrite = FALSE

fileTrain <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
##download.file(fileTrain,dest="training.csv")
buildData <-read.csv("training.csv")

set.seed(3433)
inTrain <- createDataPartition(y=buildData$classe,p=0.5,list=FALSE)
training <- buildData[inTrain,]
testing <- buildData[-inTrain,]

fileValidation <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
##download.file(fileTest,dest="testing.csv")
validation <-read.csv("testing.csv")

```


```{r utility_functions, echo = FALSE, message=FALSE,warning=F,results='hide'}

##      Many data columns obviously contain NA values for some column variables.
##      This function eliminates the columns with many or all NA values. 
                             
get_clean_data_frame <- function(data_frame){

        subset(data_frame, select = c("roll_belt","pitch_belt","yaw_belt",
        "total_accel_belt", 
        "gyros_belt_x", "gyros_belt_y","gyros_belt_z",           
        "accel_belt_x", "accel_belt_y", "accel_belt_z",
        "magnet_belt_x", "magnet_belt_y", "magnet_belt_z",
        "roll_arm", "pitch_arm", "yaw_arm",                 
        "total_accel_arm", 
        "gyros_arm_x", "gyros_arm_y", "gyros_arm_z",
        "accel_arm_x", "accel_arm_y", "accel_arm_z",
        "magnet_arm_x", "magnet_arm_y", "magnet_arm_z",
        "roll_dumbbell", "pitch_dumbbell", "yaw_dumbbell",
        "total_accel_dumbbell",    
        "gyros_dumbbell_x", "gyros_dumbbell_y", "gyros_dumbbell_z",
        "accel_dumbbell_x", "accel_dumbbell_y", "accel_dumbbell_z",
        "magnet_dumbbell_x", "magnet_dumbbell_y", "magnet_dumbbell_z",

        "roll_forearm", "pitch_forearm", "yaw_forearm",
        "total_accel_forearm", 
        "gyros_forearm_x", "gyros_forearm_y", "gyros_forearm_z",       
        "accel_forearm_x", "accel_forearm_y", "accel_forearm_z",      
        "magnet_forearm_x", "magnet_forearm_y", "magnet_forearm_z", "classe"))
}

##      The paper on the HAR study by Velloso et al. indicates that they used
##      seventeen variables and calculated statistics to predict the same
##      exercise classifications that we are concerned with. In contrast, We
##      only use the stated data in the file for our predictions without any 
##      statistics, but we are influenced by their list of seventeen 
##      parameters. We reduced the number of clean data columns from 60 to 36
##      keeping many columns of data that pertained to list of seventeen.
##      Here is a list of the seventeen critical critical parameters used by
##      Velloso et al.

##      Belt Data:
##     mean and variance of the roll
##     maximum, range and variance of the accelerometer vector,
##     variance of the gyro and variance of the magnetometer.

##     Arm:
##     variance  of  the  accelerometer  vector
##     maximum and minimum of the magnetometer

##     Dumbbell:
##     maximum of the  acceleration
##     variance  of  the  gyro
##     maximum  and minimum of the magnetometer

##     Glove:
##     sum of  the  pitch
##     maximum  and  minimum  of  the  gyro
##     With the author's suggestion as a guide we eliminate more columns from
##      consideration.


get_smaller_data_frame <- function(dat){

subset(dat, select = c("roll_belt",  
        "total_accel_belt", "gyros_belt_x", "gyros_belt_y", "gyros_belt_z",           
        "accel_belt_x", "accel_belt_y", "accel_belt_z",
        "magnet_belt_x", "magnet_belt_y", "magnet_belt_z",
        "total_accel_arm", "gyros_arm_y", "gyros_arm_z",
        "accel_arm_x", "accel_arm_y", "accel_arm_z",
        "magnet_arm_x", "magnet_arm_y", "magnet_arm_z",
        
        "total_accel_dumbbell",
        "gyros_dumbbell_x", "gyros_dumbbell_y", "gyros_dumbbell_z",
        "accel_dumbbell_x", "accel_dumbbell_y", "accel_dumbbell_z",
        "magnet_dumbbell_x", "magnet_dumbbell_y", "magnet_dumbbell_z",

        "pitch_forearm", 
        "gyros_forearm_x","gyros_forearm_y","gyros_forearm_z",       
        "magnet_forearm_x","magnet_forearm_y","magnet_forearm_z","classe"))
}

###############################################################################
###############################################################################
    
get_trained_modfit <- function(data){
                             
train(classe ~ .,
        
        roll_belt +
        total_accel_belt +
        ##gyros_belt_x +
        ##gyros_belt_y +
        gyros_belt_z +           
        ##accel_belt_x +
        ##accel_belt_y +
        ##accel_belt_z +
        ##magnet_belt_x +
        ##magnet_belt_y +
        magnet_belt_z +

    ##  arm observations:
           
        total_accel_arm +
        
        ##accel_arm_x +             
        ##accel_arm_y +
        ##accel_arm_z +
        ##magnet_arm_x +
        ##magnet_arm_y +
        magnet_arm_z +
        
   ##   dumbbell observations:

        total_accel_dumbbell +
        ##gyros_dumbbell_x +
        ##gyros_dumbbell_y +
        gyros_dumbbell_z +
        ##accel_dumbbell_x +
        ##accel_dumbbell_y +      
        ##accel_dumbbell_z +
        ##magnet_dumbbell_x +
        ##magnet_dumbbell_y +     
        magnet_dumbbell_z +

    ##  forearm (glove) observations:

        pitch_forearm +  
        ##gyros_forearm_x +
        ##gyros_forearm_y +
        gyros_forearm_z,       
        method="rf",data=data)
}

## Eliminate unused columns of data 

clean <- get_clean_data_frame(training)
train <- get_smaller_data_frame(clean)

clean <- get_clean_data_frame(testing)
test <- get_smaller_data_frame(clean)

## test <- validation

## model building

modrf <-  train(classe ~ roll_belt + total_accel_belt +
        gyros_belt_z + magnet_belt_z + total_accel_arm +
        magnet_arm_z + total_accel_dumbbell + gyros_dumbbell_z +
        magnet_dumbbell_z + pitch_forearm + gyros_forearm_z +       
        magnet_forearm_z, method="rf",data=train)
predrf <- predict(modrf,newdata=test)
trf <- table(test$classe,predrf)

accur_rf <- sum(predrf==test$classe) / length(test$classe)
accur_rf <- format(accur_rf, width = 3, digits = 3)

fitrpart <- train(classe ~ ., method = "rpart",data=train)

finMod <- fitrpart$finalModel
predrpart <- predict(fitrpart,newdata=testing)

length(predrpart)
trpart <- table(test$classe,predrpart)

accur_rpart <- sum(predrpart==test$classe) / length(test$classe)
accur_rpart <- format(accur_rpart, width = 3, digits = 3)

rf <- c(accur_rf,"Best")
rpart <- c(accur_rpart,"Runner-Up")
accuracy <- data.frame(rf,rpart)

```
                         
\newpage

#      Sample Error 

###     The accuracy of the rf method was 97.2% and the accuracy of the rpart
###     method was only 48.3%. Figure-1 below shows the confusion matrix for
###     the prediction outcome.

###     <br><br>

###     Figure-1 below shows the accuracy for prediction outcomes.

###     <br><br>

###                     Figure-1 - Accuracy

```{r accuracy_table, results='hide'}

grid.table(accuracy)


```

\newpage

```{r process_confusion_matrix, echo=FALSE,results='hide'}

## prepare table for confusion matrix

df_rf <- data.frame(trf[,1],trf[,2],trf[,3],trf[,4],trf[,5])
colnames(df_rf) <- c("A","B","C","D","E")
res.small <- xtable(df_rf)

```

#      Confusion Matrix 

###     Figure-2 below shows the confusion matrix for prediction outcomes.

###     <br><br>

###                     Figure-2 - Confusion Matrix
```{r confusion_matrix, results='asis'}

grid.table(df_rf)


```

##      Quiz Trial

###     <br><br>

###     The test cases were evaluated by setting the test data for the quiz
###     equal to a validation data frame. Before the final evaluation the
###     validation data was substituted for the training data that was used
###     for model building.
```{r quiz, echo = FALSE, message=FALSE, warning=F,results='hide'}

## test <- validation

## model building

predrf <- predict(modrf,newdata=validation)

Question <- c(1:20)
Exercise_Grade <- predrf

Test_Answers <- data.frame(Question,Exercise_Grade)

``` 

###     <br><br>

###                     Figure-3 - Trial Run for Quiz Anwsers
```{r models, results='asis', message=F, warning=F, fig.width=5, fig.height=6}

grid.table(Test_Answers)
        
```

###     <br><br>

#       Reference

###     <br><br>
###     <br><br>

###     Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H.
###     Qualitative Activity Recognition of Weight Lifting Exercises.
###     Proceedings of 4th International Conference in Cooperation with SIGCHI
###     (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.




